---
title: "Introduction to the surveydata package."
author: "Andrie de Vries"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the surveydata package.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


`surveydata` is a package that makes it easy to work with typical survey data that originated in SPSS or other formats.

A surveydata object consists of:

* A data frame with a row for each respondent and a column for each question.   Column names are typically names in the pattern `Q1`, `Q2_1`, `Q2_2`, `Q3` - where underscores separate the subquestions when these originated in a grid (array) of questions.

* Question metadata gets stored in the `{variable.labels} attribute of the data frame. This typically contains the original questionnaire text for each question.
  
* Information about the subquestion separator (typically an underscore) is stored in the `patterns` attribute.
  

Data processing a survey file can be tricky, since the standard methods for dealing with data frames does not conserve the `variable.labels` attribute.  The `surveydata` package defines a `surveydata` class and the following methods that knows how to deal with the `variable.labels` attribute:

* `as.surveydata`
* `[.surveydata`
* `[<-.surveydata`
* `\$.surveydata`
* `\$<-.surveydata`
* `merge.surveydata`

In addition, `surveydata` defines the following convenient methods for extracting and working with the variable labels:

* `varlabels`
* `varlabels<-`


## Defining a surveydata object

First load the `surveydata` package.

```{r Setup}
library(surveydata)
```

Next, create sample data. A data frame is the ideal data structure for survey data, and the convention is that data for each respondent is stored in the rows, while each column represents answers to a specific question.

```{r sample-data}

sdat <- data.frame(
    id   = 1:4,
    Q1   = c("Yes", "No", "Yes", "Yes"),
    Q4_1 = c(1, 2, 1, 2), 
    Q4_2 = c(3, 4, 4, 3), 
    Q4_3 = c(5, 5, 6, 6), 
    Q10 = factor(c("Male", "Female", "Female", "Male")),
    crossbreak  = c("A", "A", "B", "B"), 
    weight      = c(0.9, 1.1, 0.8, 1.2)
)

```


The survey metadata consists of the questionnaire text.  For example, this can be represented by a character vector, with an element for each question.

To assign this metadata to the survey data, use the `varlabels()` function.  This function assigns the questionnaire text to the `variable.labels` attribute of the data frame.

```{r varlabels}

varlabels(sdat) <- c(
    "RespID",
    "Question 1", 
    "Question 4: red", "Question 4: green", "Question 4: blue", 
    "Question 10",
    "crossbreak",
    "weight"
  )
```


Finally, create the surveydata object. To do this, call the `as.surveydata()` function.  The argument `renameVarlabels` controls whether the `varlabels` get renamed with the same names as the data.  This is an essential step, and ensures that the question text remains in synch with the column names.

```{r init}
sv <- as.surveydata(sdat, renameVarlabels = TRUE)
```


## Extracting specific questions

It is easy to extract specific questions with the `[` operator. This works very similar to extraction of data frames. However, there are two important differences:

* The extraction operators will always return a `surveydata` object, even if only a single column is returned. This is different from the behaviour of data frames, where a single column is simplified to a vector.
* Extracing a question with multiple subquestions, e.g. "Q4" returns multiple columns


```{r extract}
sv[, "Q1"]
sv[, "Q4"]
```

The extraction makes use of the underlying metadata, contained in the `varlabels` and `pattern` attributes:

```{r attributes}

varlabels(sv)
pattern(sv)
```

## Working with question columns

It is easy to query the surveydata object to find out which questions it contains, as well as which columns store the data for those questions.

```{r questions}
questions(sv)
which.q(sv, "Q1")
which.q(sv, "Q4")
```

## Reading the questionnaire text

The function `qText()` gives access to the questionnaire text.

```{r qtext}
qText(sv, "Q1")
qText(sv, "Q4")

qTextCommon(sv, "Q4")
qTextUnique(sv, "Q4")
```
